\ourTitle

### 2022-06-06 paper/manuscript.tex
JPC: Started editing.

%%
%% This is file `sample-lualatex.tex', % generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed % with new filenames
%distinct from sample-lualatex.tex.
%% 
%% For distribution of the original source see the terms % for copying and
%modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the % original source
%files, as listed above, are part of the % same distribution. (The sources need
%not necessarily be % in the same archive or directory.)
%%
%%
%% Commands for TeXCount TC:macro \cite [option:text,text] TC:macro \citep
%[option:text,text] TC:macro \citet [option:text,text] TC:envir table 0 1
%TC:envir table* 0 1 TC:envir tabular [ignore] word TC:envir displaymath 0 word
%TC:envir math 0 word TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[draft,aps,pre,twocolumn,groupaddress,showkeys,nofootinbib,preprintnumbers,floatfix]{revtex4-2}
\usepackage{probs}
\usepackage{dynlearn}
\usepackage[linesnumbered,ruled]{algorithm2e}
\mathchardef\mhyphen="2D

@@

\preprint{\arxiv{2108.XXXXX}}

@@

\tableofcontents
% Clears up font spacing/overlap in ToC
% with sufficient repetitions of formating (e.g., pdflatex).
% Other front matter packages can alter formating constants.
\makeatletter
\let\toc@pre\relax
\let\toc@post\relax
\makeatother

@@

\section{Introduction}
Suppose that we have a finite sequence $x_1 \dots x_L$ of categorical
observations drawn from a temporal process. We may suppose that the process is
stationary (time-translation invariant) and ergodic (explores
all possible behaviors) \cite{Kall01a,Shal07a}. We may wish to forecast from the
observed information the behavior of the next $n$ observations. If we suspect
that the process' temporal correlations do not matter much beyond $k$ symbols,
then we take our data to be all subsequences of length $n+k$,
splitting each subsequence into words of length $k$ and $n$ respectively. From
these ``past/future'' pairs we can construct an empirical conditional
distribution
\begin{align*}
  \widehat{P}_{x_{1}\dots x_{n}|x_{-k+1}\dots x_0} = 
  \frac{C_{x_{-k+1}\dots x_{n}}}{C_{x_{-k+1}\dots x_0}}
  ~,
\end{align*}
where $C_{w}$ is the number of times the word $w$ appears in our sequence $x_1
\dots x_L$.

@@

Predictive states are employed for inference and modeling in dynamical
systems \cite{Crut92c,Emen16a,Salo19a}, renewal processes and spike-trains
\cite{Marz14e,Marz15a}, condensed matter physics \cite{Varn12a,Varn14a}, and
spatiotemporal systems \cite{Rupe19a,Rupe19b,Rupe20a}. A deep mathematical
theory of predictive-state inference has been correspondingly developed
\cite{Uppe97a,Jaeg00a,Shal02a,Jame04a,Stil07b,Stre13a,Thon15a,Marz19a,Brod20a}.
If the dataset $x_1 \dots x_L$ is drawn from any
stationary process and if $L$ is sufficiently large, then Eq.
\eqref{eq:convergence} converges for any $n$ and a probability-$1$ subset of
pasts $\overleftarrow{x}$ \cite{Uppe97a}. Recently, this result has been further
clarified: in the language of measures,
$P_{\overleftarrow{x}}$ converges \emph{in distribution}, with respect to the
\emph{product topology} of the space of sequences $\mathcal{X}^\mathbb{N}$
\cite{Loom21a}.

@@

The following focuses on the first, as it is a crucial building
block to achieving the other two. Recent attempts to reconstruct the geometry of predictive states embedded them
in reproducing kernel Hilbert spaces
\cite{Song09a,Song10a,Boot13a,Brod20a,Loom21a}. This was achieved to great
effect largely since convergence in Hilbert spaces generated by universal
kernels is equivalent to convergence in distribution \cite{Srip10a}. That is,
these embeddings allow accurate representations of predictive states because
they respect the product topology of sequences in the same way that predictive
states themselves do \cite{Loom21a}. In this paper, we will seek to achieve the same results,
using a procedure which can be easily visualized and interpreted, making it
suitable as a method for exploratory data analysis \cite{Tuke62a,Tuke77a}.

@@

We refer back frequently to the following example processes:
\begin{enumerate}
\item The \emph{even process} can be generated by repeatedly tossing a coin
	and writing down a $0$ for every tail and $11$ for every head. The process
	is essentially random except that $1$s only appear in contiguous blocks of
	even size bounded by $0$s. The even process has infinite Markov order but
	can be generated by a two-state hidden Markov chain \cite{Crut01a}. A typical
	example might look like $01100111101100011$.
\item The $a^n b^n$ \emph{process} can be generated by choosing a random
	integer $n\geq 1$ (we suppose via a Poisson process) and writing $n$
	$\mathtt{a}$'s followed by an equal number of $\mathtt{b}$'s, and then
	repeating this procedure indefinitely. This results in sequences where any
	contiguous block of $\mathtt{a}$s is followed by a block of $\mathtt{b}$s
	of equal size. The $\mathtt{a}^n \mathtt{b}^n$ process cannot be generated
	by any finite hidden Markov chain, though it is a simple example of a
	probabilistic context-free language \cite{Hopc06a}. A typical example might
	look like $\mathtt{abaaabbbabaabb}$.
\item The $x+f(x)$ \emph{process} is a probabilistic context-free language
	modeling the syntactic structure of simple mathematical expressions. It has
	terminal symbols $\{\mathtt{(}\ ,\ \mathtt{)}\ ,\ \mathtt{;}\ ,\
	\mathtt{+}\ ,\ \mathtt{f}\ ,\ \mathtt{x}\}$ and nonterminals $\{A,B,C\}$,
	and starts with a sequence of $A$s. Sequences are generated by applying the
	production rules:
\begin{align*}
    A &\mapsto B\ \mathtt{+}\ C\ \mathtt{;}\ |\ C\ \mathtt{;}\\
    B &\mapsto B\ \mathtt{+}\ C\ |\ C\\
    C &\mapsto \mathtt{f(}B\mathtt{)}\ |\ \mathtt{x}
  ~.
\end{align*}
A typical example might
	look like $\mathtt{x+x;f(x+f(x));f(f(x));}$.
\item The $a^n b^n c^n$ \emph{process} is a probabilistic indexed language
	\cite{Hopc06a} that is analogous to $\mathtt{a}^n \mathtt{b}^n$ except
	after writing the blocks of $\mathtt{a}$'s and $\mathtt{b}$'s, we also
	write a block of $\mathtt{c}$'s of length $n$. A typical example might
	look like $\mathtt{abcaaabbbcccabcaabbcc}$.
\item The \emph{Morse-Thue process} is generated by sampling from the time
	series of the logistic map at critical ``onset of chaos'' parameter
	$r_c \approx 3.56995$:
\begin{align*}
    y_{t+1} = r y_t(1-y_t)
\end{align*}
	and then coarse-graining the data by taking $x_t = 0$ if $0< y_t \leq
	\frac{1}{2}$ and $x_t = 1$ if $\frac{1}{2} < y_t <1$ \cite{Kurk03a}.
	Alternatively, we can generate this process by starting with a single $0$
	and executing the replacements $0 \mapsto 11$ and $1 \mapsto 01$
	consecutively. The resulting process is an indexed-context free language
	\cite{Crut92c}. A typical example might look like
	$11011101010111011101110101011101$ (this is the fifth generation of the
	replacement rule starting from $0$).
\end{enumerate}

@@

@@

      \eIf{$\overleftarrow{\mathbf{x}} \in \UnqPasts$} { 
        $\Append\ \overleftarrow{\mathbf{x}}$ \KwTo $\UnqPasts$\; 
        $\Append\ [p]$ \KwTo $\Cantors$\; 
        } { 
        $j \leftarrow \Index\left(\overleftarrow{\mathbf{x}},\UnqPasts\right)$\;
        $\Append\ p$ \KwTo $\Cantors_j$\; 
        } 
  }
  $K \leftarrow \Length(\UnqPasts)$\; 
  $\Wass \leftarrow \Matrix(K,K)$\;
  \For{$i\leftarrow 1$ \KwTo $K$} { 
    \For{$j\leftarrow 1$ \KwTo $K$} {  
      $\Wass_{ij}
        \leftarrow
        \Wasserstein(\Cantors_i,\Cantors_j)$\;
        $\Wass_{j i} \leftarrow \Wass_{i j}$\; } } 
  \KwResult{$\UnqPasts$,$\Cantors$,$\Wass$}
  \caption{Steps for converting a sequence of categorical time-series data
        into a labeled collection of empirical distributions of Cantor-embedded futures,
        and a matrix of Wasserstein distances from said distributions.}
  \label{alg:cantorwass}
\end{algorithm}

>>>>

% Exploring Predictive States via Cantor Embeddings and Wasserstein Distance
%
% spl: 6/2/22
% jpc: 6/5/22

%% Commands for TeXCount TC:macro \cite [option:text,text] TC:macro \citep
%[option:text,text] TC:macro \citet [option:text,text] TC:envir table 0 1
%TC:envir table* 0 1 TC:envir tabular [ignore] word TC:envir displaymath 0 word
%TC:envir math 0 word TC:envir comment 0 0
%%

\documentclass[draft,aps,pre,twocolumn,groupaddress,showkeys,nofootinbib,preprintnumbers,floatfix]{revtex4-2}
\usepackage{probs}
\usepackage{dynlearn}
\usepackage[linesnumbered,ruled]{algorithm2e}
\mathchardef\mhyphen="2D

@@

\preprint{\arxiv{2206.XXXXX}}

@@

% \tableofcontents
% Clears up font spacing/overlap in ToC
% with sufficient repetitions of formating (e.g., pdflatex).
% Other front matter packages can alter formating constants.
\makeatletter
\let\toc@pre\relax
\let\toc@post\relax
\makeatother

@@

\section{Introduction}
Suppose that we have a finite sequence $x_1 \dots x_L$ of categorical
observations drawn from a temporal process. We may suppose that the process is
stationary (time-translation invariant) and ergodic (explores
all possible behaviors) \cite{Fell70a,Kall01a}. We may wish to forecast from the
observed information the behavior of the next $n$ observations. If we suspect
that the process' temporal correlations do not matter much beyond $k$ symbols,
then we take our data to be all subsequences of length $n+k$,
splitting each subsequence into words of length $k$ and $n$ respectively. From
these ``past/future'' pairs we can construct an empirical conditional
distribution
\begin{align*}
  \widehat{P}_{x_{1}\dots x_{n}|x_{-k+1}\dots x_0} = 
  \frac{C_{x_{-k+1}\dots x_{n}}}{C_{x_{-k+1}\dots x_0}}
  ~,
\end{align*}
where $C_{w}$ is the number of times the word $w$ appears in our sequence $x_1
\dots x_L$.

@@

Predictive states are employed for inference and modeling in dynamical systems
\cite{Crut92c}, renewal processes and neural spike-trains
\cite{Marz14e,Marz15a}, condensed matter physics \cite{Varn14a}, and
spatiotemporal systems \cite{Rupe19b}. A deep mathematical
theory of predictive-state inference has been correspondingly developed
\cite{Uppe97a,Jaeg00a,Shal02a,Jame04a,Stil07b,Stre13a,Thon15a,Brod20a}.
If the dataset $x_1 \dots x_L$ is drawn from any stationary process and if $L$
is sufficiently large, then Eq. \eqref{eq:convergence} converges for any $n$
and a probability-$1$ subset of pasts $\overleftarrow{x}$ \cite{Uppe97a}.
Recently, this result has been further clarified: in the language of measures,
$P_{\overleftarrow{x}}$ converges \emph{in distribution}, with respect to the
\emph{product topology} of the space of sequences $\mathcal{X}^\mathbb{N}$
\cite{Loom21a}.

@@

The following focuses on the first, as it is a crucial building block to
achieving the other two. Recent attempts to reconstruct the geometry of
predictive states embedded them in reproducing kernel Hilbert spaces
\cite{Song09a,Song10a,Boot13a,Brod20a,Loom21a}. This was achieved to great
effect largely since convergence in Hilbert spaces generated by universal
kernels is equivalent to convergence in distribution \cite{Srip10a}. That is,
these embeddings allow accurate representations of predictive states because
they respect the product topology of sequences in the same way that predictive
states themselves do \cite{Loom21a}. Here, we achieve the same results using a
procedure that can be easily visualized and interpreted. This makes it suitable
as a method for exploratory data analysis \cite{Tuke62a,Tuke77a}, while moving
further toward interpretable machine learning of structured processes.

@@

We refer back frequently to the following example processes of increasing
computational complexity:
\begin{enumerate}
      \setlength{\topsep}{-2pt}
      \setlength{\itemsep}{-2pt}
      \setlength{\parsep}{-2pt}
\item The \emph{even process} can be generated by repeatedly tossing a coin
	and writing down a $0$ for every tail and $11$ for every head. The process
	is essentially random except that $1$s only appear in contiguous blocks of
	even size bounded by $0$s. The even process has infinite Markov order but
	can be generated by a two-state hidden Markov chain \cite{Crut01a}. A
	typical example might look like $01100111101100011$.
\item The $\mathtt{a}^n \mathtt{b}^n$ \emph{process} can be generated by
	choosing a random integer $n\geq 1$ (we suppose via a Poisson process) and
	writing $n$ $\mathtt{a}$s followed by an equal number of $\mathtt{b}$s,
	and then repeating this procedure indefinitely. This results in sequences
	where any contiguous block of $\mathtt{a}$s is followed by a block of
	$\mathtt{b}$s of equal size. The $\mathtt{a}^n \mathtt{b}^n$ process cannot
	be generated by any finite hidden Markov chain, though it is a simple
	example of a probabilistic context-free language \cite{Hopc06a}. A typical
	example might look like $\mathtt{abaaabbbabaabb}$.
\item The $x+f(x)$ \emph{process} is a probabilistic context-free language
	modeling the syntactic structure of simple mathematical expressions. It has
	terminal symbols $\{\mathtt{(}\ ,\ \mathtt{)}\ ,\ \mathtt{;}\ ,\
	\mathtt{+}\ ,\ \mathtt{f}\ ,\ \mathtt{x}\}$ and nonterminals $\{A,B,C\}$,
	and starts with a sequence of $A$s. Sequences are generated by applying the
	production rules:
\begin{align*}
    A &\mapsto B\ \mathtt{+}\ C\ \mathtt{;}\ |\ C\ \mathtt{;}\\
    B &\mapsto B\ \mathtt{+}\ C\ |\ C\\
    C &\mapsto \mathtt{f(}B\mathtt{)}\ |\ \mathtt{x}
  ~.
\end{align*}
	A typical example might look like $\mathtt{x+x;f(x+f(x));f(f(x));}$.
\item The $\mathtt{a}^n \mathtt{b}^n \mathtt{c}^n$ \emph{process} is a
	probabilistic indexed language \cite{Hopc06a} that is analogous to
	$\mathtt{a}^n \mathtt{b}^n$ except after writing the blocks of
	$\mathtt{a}$'s and $\mathtt{b}$'s, we also write a block of $\mathtt{c}$'s
	of length $n$. A typical example might look like
	$\mathtt{abcaaabbbcccabcaabbcc}$.
\item The \emph{Morse-Thue process} is generated by sampling from the time
	series of the logistic map at critical ``onset of chaos'' parameter
	$r_c \approx 3.56995$:
\begin{align*}
    y_{t+1} = r y_t(1-y_t)
\end{align*}
	and then coarse-graining the data by taking $x_t = 0$ if $0< y_t \leq
	\frac{1}{2}$ and $x_t = 1$ if $\frac{1}{2} < y_t <1$ \cite{Kurk03a}.
	Alternatively, we can generate this process by starting with a single $0$
	and executing the replacements $0 \mapsto 11$ and $1 \mapsto 01$
	consecutively. The resulting process is an indexed-context free language
	\cite{Crut92c}. A typical example might look like
	$11011101010111011101110101011101$---the fifth generation of the
	replacement rule starting from $0$.
\end{enumerate}

@@

STOPPED

@@

      \eIf{$\overleftarrow{\mathbf{x}} \in \UnqPasts$} { 
        $\Append\ \overleftarrow{\mathbf{x}}$ \KwTo $\UnqPasts$\; 
        $\Append\ [p]$ \KwTo $\Cantors$\; 
        } { 
        $j \leftarrow \Index\left(\overleftarrow{\mathbf{x}},\UnqPasts\right)$\;
        $\Append\ p$ \KwTo $\Cantors_j$\; 
        } 
  }
  $K \leftarrow \Length(\UnqPasts)$\; 
  $\Wass \leftarrow \Matrix(K,K)$\;
  \For{$i\leftarrow 1$ \KwTo $K$} { 
    \For{$j\leftarrow 1$ \KwTo $K$} {  
      $\Wass_{ij}
        \leftarrow
        \Wasserstein(\Cantors_i,\Cantors_j)$\;
        $\Wass_{j i} \leftarrow \Wass_{i j}$\; } } 
  \KwResult{$\UnqPasts$,$\Cantors$,$\Wass$}
\caption{Convert a sequence of categorical time-series data into a labeled
	collection of empirical distributions of Cantor-embedded futures and a
	matrix of Wasserstein distances between said distributions.
	}
\label{alg:cantorwass}
\end{algorithm}
